---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
author: Jonathan Hines
keep_md: true
---
```{r echo=FALSE}
## Download the data if not present
# If the source isn't downloaded yet go get it and download it
if (!file.exists("activity.csv")) {
    rawDataSource = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip";
    if( .Platform$OS.type == "windows" ) {
        download.file(rawDataSource, destfile = "activity.zip", mode="wb")
        unzip("activity.zip")
        unlink("activity.zip")
    } else {
        download.file(rawDataSource, destfile = "activity.zip", method="curl")
        unzip("activity.zip")
        unlink("activity.zip")
    }
    write(date(), file="dateDownloaded.txt")
    rm( rawDataSource )
}
```

## Loading and preprocessing the data
```{r}
a <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?
Generate a vector of the number of steps taken in each day in the data set
```{r}
steps_per_day <- tapply(a$steps, a$date, sum, na.rm = TRUE)
```

Here is a histogram of the number of steps taken on a given day
```{r}
hist(
    steps_per_day,
    main="Histogram of the number of steps taken in a day", 
    xlab="Steps in a day"
)
```

The mean number of steps per day is given by:
```{r}
mean(steps_per_day)
```

The median number of steps is given by:
```{r}
median(steps_per_day)
```

## What is the average daily activity pattern?
Generate a vector of the average number of steps split by interval
```{r}
mean_steps_per_interval <- tapply(a$steps, a$interval, mean, na.rm = TRUE)
steps_intervals <- strtoi(attr(mean_steps_per_interval, "dimnames")[[1]])
```

Plot a time series 
```{r}
plot(
    steps_intervals, 
    mean_steps_per_interval, 
    type="l", 
    main="Average number of steps over a 5 minute interval", 
    xlab="Interval", 
    ylab="Mean number of steps per interval"
)
```

The interval that contains the hightest number of steps on average can be
determined by:
```{r}
steps_intervals[which.max(mean_steps_per_interval)]
```

## Inputing missing values
The total number of missing values in the data set is:
```{r}
sum(is.na(a$steps))
```

To deal with the missing values, replace the median for that interval.  There
is definitely a faster way to do this than with a for loop but this one works
fine for now
```{r}
median_steps_per_interval <- tapply(a$steps, a$interval, median, na.rm = TRUE)
a_na_replace <- a
for(i in 1:nrow(a_na_replace)) {
    if(is.na(a_na_replace[i,"steps"])) {
        a_na_replace[i,"steps"] <- median_steps_per_interval[
            which(steps_intervals==a_na_replace[i,"interval"])
        ]
    }
}
```

Here is a histogram of the step data with the NAs replaced by the medians of
the corresponding interval.

```{r}
steps_per_day_na_replace <- tapply(
    a_na_replace$steps,
    a_na_replace$date,
    sum
)
hist(
    steps_per_day_na_replace,
    main="Histogram of the number of steps taken in a day\nwith NA's replaced with medians", 
    xlab="Steps in a day"
)
```
It appears that adding in missing data has taken days that recorded the
lowest number of steps and increased their values.  This makes sense considering
that we are effectivly adding steps to intervals that did not previously 
contribute to the number of steps in their respective day.

The mean number of steps per day is found by:
```{r}
mean(steps_per_day_na_replace)
```

The median number of steps taken per day is found by:
```{r}
median(steps_per_day_na_replace)
```

As a reminder here is the mean and median values respectively, with the
missing values counted as 0:
```{r}
mean(steps_per_day)
median(steps_per_day)
```

It appears the the missing data on the estimates of total numbers of steps
per day is minor or `r mean(steps_per_day_na_replace)/mean(steps_per_day)`% for
the mean and `r median(steps_per_day_na_replace)/median(steps_per_day)`% for the
median.

## Are there differences in activity patterns between weekdays and weekends?
Make a new column in the filled in dataset that identifies the date as a 
weekday.

```{r}
a_weekdays <- weekdays(as.Date(a_na_replace$date))

```
